
**Loading Data**  
setting the directory:
```{r setdir, cache=TRUE}
      currentDirectory <- getwd()
      newDirectory <- paste(currentDirectory,"/PeerAssessment1",sep="")
      if(!file.exists("PeerAssessment1")){dir.create("PeerAssessment1")}
      setwd(newDirectory)
```

downloading, unzipping and reading the data:
```{r loaddata, cache=TRUE}
      setInternet2()
      file_train <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
      download.file(file_train,"file_train.csv")
      train_data <- read.csv("file_train.csv") ## data for model generation
```

**Loading Packages**
```{r packages, cache=FALSE, message=FALSE, warning=FALSE}
      library(caret)
      library(kernlab)
      library(gbm)
      library(randomForest)
      library(RRF)
      library(e1071)
      library(kernlab)
      library(nnet)
      library(xtable)
```

**Tagging of Colmns**
For the model generation and prediction only columns with less no more than 1000 NA entries are selected and for the PCA only columns with numeric values can be used. The selecting of those columns is performed in the next chunk.
```{r dataclean, cache=TRUE}
      train_data <- train_data[,-1]

      train_data <- transform(train_data,"hour"=as.factor(strptime(train_data$cvtd_timestamp,
                                                                   format="%d/%m/%Y %H:%M")$hour))
      train_data <- transform(train_data,"year"=as.factor(as.Date(train_data$cvtd_timestamp,
                                                                     format="%d/%m/%Y")))
      
      train_data <- train_data[,!grepl("time",colnames(train_data))]  

      train_data <- train_data[,!grepl("window",colnames(train_data))]



    ## filtering NA columns
      na_count <- sapply(train_data,function(col){
                      sum(is.na(col))})
      na_count2 <- sapply(train_data,function(col){
                      is.element("#DIV/0!",col)})
      na_count  <- na_count > 1000

    
      ##classes <- sapply(train_data,class)
      ##for(i in colnames(train_data)){
      ##        if(is.integer(train_data[,i])|
      ##           is.logical(train_data[,i])){
      ##          train_data[,i]<- as.numeric(train_data[,i])}
      ##        }
    ## filtering factor columns
      factorVar <- sapply(train_data,is.factor)
      ##DateVar <- grepl("year",colnames(train_data))
```

**Generation of test sets**
creating training, testing and validation set
```{r gensets}
      inTrain   <- createDataPartition(y=train_data$classe,p=0.6,list=F)
      train_set <- train_data[inTrain,] ## train set
      no_train_set <- train_data[-inTrain,]
      inTest    <- createDataPartition(y=no_train_set$classe,p=0.5,list=F)
      test_set  <- no_train_set[inTest,] ## test set
      validation_set <-  no_train_set[-inTest,] ## validation set
```


**Principal Component Analysis**
Generation of PC with train set and computation of PCA-values with training, testing and validation set data. PCA is only based on columns with numeric values and those with less than 1000 NA entries (see previous code chunk).
```{r pca, cache=TRUE}
    ## Construction 15 principal components      
      PCAtrain <- preProcess(train_set[,!na_count2&!na_count&!factorVar],
                                          method="pca",pcaComp=20)

    ## Computation of PCA for training testing and validation data set
    ## Since PCA is only performed based on numeric columns factor columns are attached 
    ## to generated PC after there generation
      PCAtrain_val <- predict(PCAtrain,train_set[,!na_count2&!na_count&!factorVar])
      train_factor_var <- train_set[,colnames(train_set)[!na_count2&!na_count&factorVar]]
      ##train_var <- cbind(train_factor_var[,c("user_name","classe")],PCAtrain_val)
      train_var <- cbind(train_factor_var,PCAtrain_val)
      ##[,c("user_name","classe")]
      PCAtest_val <- predict(PCAtrain,test_set[,!na_count2&!na_count&!factorVar])
      test_factor_var <- test_set[,colnames(train_set)[!na_count2&!na_count&factorVar]]
      ##test_var <- cbind(test_factor_var[,c("user_name","classe")],PCAtest_val)      
      test_var <- cbind(test_factor_var,PCAtest_val)   
      
      PCAvalidation_val <- predict(PCAtrain,validation_set[,!na_count2&!na_count&!factorVar])
      validation_factor_var <- validation_set[,colnames(train_set)[!na_count2&!na_count&factorVar]]
      ##validation_var <- cbind(validation_factor_var[,c("user_name","classe")],PCAvalidation_val) 
      validation_var <- cbind(validation_factor_var,PCAvalidation_val) 
```

### Individual Model Generation and Individual Model Prediction
In the following the six model considered are generated - gradient boosed machines, random forest, randomized random forest, naive bayes, support vector machines and neural networks - using the training set and the corresponding prediction for training, testing and validation sets are made.

**Gradient Bosted Machines (gbm)**
```{r gbm, cache=TRUE}
    ## gbm model generation 
      modelFit_gbm <- gbm(classe~., data=train_var, 
                n.trees = 1000, shrinkage = 0.01, 
                interaction.depth = 4,
                distribution = "multinomial")
    ## training set prediction and accuracy
      gbm_pred_train <- predict(modelFit_gbm, train_var, n.trees=1000, type="response")
      ## The predictions does not contain the factor values but are numeric, 
      ## wherefore they need to be transformed with the following code 
      gbm_train <- confusionMatrix(train_set$classe,factor(colnames(gbm_pred_train
                                    )[max.col(data.frame(gbm_pred_train))],levels=c("A","B","C","D","E")))            
    ## testing set prediction
      gbm_pred_test <- predict(modelFit_gbm, test_var, n.trees=1000, type="response")
      gbm_test <- confusionMatrix(test_set$classe,factor(colnames(gbm_pred_test
                                    )[max.col(data.frame(gbm_pred_test))],levels=c("A","B","C","D","E")))  
    
    ## validation set prediction
      gbm_pred_val <- predict(modelFit_gbm, validation_var, n.trees=1000, type="response")
      gbm_val <- confusionMatrix(validation_set$classe,factor(colnames(gbm_pred_val
                                    )[max.col(data.frame(gbm_pred_val))],levels=c("A","B","C","D","E")))  

```

**Random Forest (rf)**
```{r rf, cache=TRUE}
    ## rf model generation
      modelFit_rf <- randomForest(classe~., data=train_var)
    ## training testing and validation set predictions and accuracy
      rf_train <- confusionMatrix(train_set$classe,predict(modelFit_rf,train_var))
      rf_test  <- confusionMatrix(test_set$classe,predict(modelFit_rf,test_var))
      rf_val  <- confusionMatrix(validation_set$classe,predict(modelFit_rf,validation_var))
```

**Randomized Random Forest (rrf)**
```{r rrf, cache=TRUE}
    ## rrf model generation
      modelFit_rrf <- RRF(classe~., data=train_var)
    ## training testing and validation set predictions and accuracy
      rrf_train <- confusionMatrix(train_set$classe,predict(modelFit_rrf,train_var))
      rrf_test  <- confusionMatrix(test_set$classe,predict(modelFit_rrf,test_var))
      rrf_val <- confusionMatrix(validation_set$classe,predict(modelFit_rrf,validation_var))
```

**Naive Bayes (nb)**
```{r nb, cache=TRUE}
    ## nb model generation
      modelFit_nb <- naiveBayes(classe~., data=train_var)
    ## training testing and validation set predictions and accuracy
      nb_train <- confusionMatrix(train_set$classe,predict(modelFit_nb,train_var))
      nb_test <- confusionMatrix(test_set$classe,predict(modelFit_nb,test_var))
      nb_val <- confusionMatrix(validation_set$classe,predict(modelFit_nb,validation_var))
```

**Support Vector Machines (svm)**
```{r svm, cache=TRUE}
    ## svm model generation
      modelFit_svm <- svm(classe~., data=train_var)
    ## training testing and validation set predictions and accuracy
      svm_train <- confusionMatrix(train_set$classe,predict(modelFit_svm,train_var))
      svm_test  <- confusionMatrix(test_set$classe,predict(modelFit_svm,test_var))
      svm_val  <- confusionMatrix(validation_set$classe,predict(modelFit_svm,validation_var))
```

**Neural Networks (nnet)**
```{r nnet, cache=TRUE, results='hide'}
    ## nnet model generation
      modelFit_nnet<- nnet(classe~., data=train_var,size=20)
    ## training testing and validation set predictions and accuracy
      ## The predictions does not contain the factor values but are numeric, 
      ## wherefore they need to be transformed with the following code 
      nnet_pred_train <- predict(modelFit_nnet,train_var)
      nnet_train <- confusionMatrix(train_set$classe,factor(colnames(nnet_pred_train
                                  )[max.col(data.frame(nnet_pred_train))],levels=c("A","B","C","D","E")))  
      
      nnet_pred_test  <- predict(modelFit_nnet,test_var)
      nnet_test <- confusionMatrix(test_set$classe,factor(colnames(nnet_pred_test
                                  )[max.col(data.frame(nnet_pred_test))],levels=c("A","B","C","D","E")))
      
      nnet_pred_val  <- predict(modelFit_nnet,validation_var)
      nnet_val <- confusionMatrix(validation_set$classe,factor(colnames(nnet_pred_val
                                  )[max.col(data.frame(nnet_pred_val))],levels=c("A","B","C","D","E"))) 
```

**Summarizing the Model's Results**
```{r sumdata, cache=TRUE}
    ## training set prediction accuracy
      train_table <- rbind("rf"=rf_train$overall,"rrf"=rrf_train$overall,"nb"=nb_train$overall,
                           "svm"=svm_train$overall,"gbm"=gbm_train$overall,"nnet"=nnet_train$overall)
      train_table <- train_table[,1:5]
    ## testing set prediction accuracy
      test_table <- rbind("rf"=rf_test$overall,"rrf"=rrf_test$overall,"nb"=nb_test$overall,
                           "svm"=svm_test$overall,"gbm"=gbm_test$overall,"nnet"=nnet_test$overall)
      test_table <- test_table[,1:5]
    ## testing set validation accuracy
      val_table <- rbind("rf"=rf_val$overall,"rrf"=rrf_val$overall,"nb"=nb_val$overall,
                         "svm"=svm_val$overall,"gbm"=gbm_val$overall,"nnet"=nnet_val$overall)
      val_table <- val_table[,1:5]
```

### Generating the Combined Predictions
In the following the gbm-, rf-, nb-,svm- and nnet-prediction models are used to generate the combined model (comb) which is combuted using the best performing algorithm (this is random forest (rf), see the results later).

**Generating Model and Prediction test-set Values**
```{r gencombpre,cache=TRUE}
    ## summarizing individual model's predictions in one data frame
      predDF_test <- data.frame("rf"=predict(modelFit_rf,test_var),"nb"=predict(modelFit_nb,test_var),
                          "svm"=predict(modelFit_svm,test_var),
                          "gbm"=factor(colnames(gbm_pred_test)[max.col(data.frame(gbm_pred_test))],levels=c("A","B","C","D","E")),
                          "nnet"=factor(colnames(nnet_pred_test)[max.col(data.frame(nnet_pred_test))],levels=c("A","B","C","D","E")),
                          "outc"=test_set$classe)
    ## generating the model
      combModFit <- randomForest(outc~., data=predDF_test)
    ## predicting values
      combPred_test <- predict(combModFit,predDF_test)
    ## computin accuracy
      comb_test <- confusionMatrix(test_set$classe,predict(combModFit,predDF_test))
```

**Predicting validation-set Values**
```{r preval,cache=TRUE}
    ## summarizing individual model's predictions in one data frame
      predDF_val <- data.frame("rf"=predict(modelFit_rf,validation_var),"nb"=predict(modelFit_nb,validation_var),
                          "svm"=predict(modelFit_svm,validation_var),
                          "gbm"=factor(colnames(gbm_pred_val)[max.col(data.frame(gbm_pred_val))],levels=c("A","B","C","D","E")),
                          "nnet"=factor(colnames(nnet_pred_val)[max.col(data.frame(nnet_pred_val))],levels=c("A","B","C","D","E")),
                          "outc"=validation_set$classe)
    ## predicting validation-set values and accuracy     
      combPred_val <- predict(combModFit,predDF_val)
      comb_val <- confusionMatrix(validation_set$classe,predict(combModFit,predDF_val))
```

**Attaching Results to the Tables**
```{r cache=TRUE}
      test_table <- rbind(test_table,"comb"=comb_test$overall[1:5])
      val_table  <- rbind(val_table,"comb"=comb_val$overall[1:5])
```

```{r results = 'asis'}
      html_attributes <- list('border="1" align="center"')
      print(xtable(train_table,caption="Accuracy Training Data from training.csv",digits=c(6,3,3,3,3,3)),
            type="html",html.table.attributes=html_attributes)
      print(xtable(test_table,caption="Accuracy Testing Data from training.csv",digits=c(6,3,3,3,3,3)),
            type="html",html.table.attributes=html_attributes)
      print(xtable(val_table,caption="Accuracy Validation Data from training.csv",digits=c(6,3,3,3,3,3)),
            type="html",html.table.attributes=html_attributes)
```

### Prediction for test.csv
**Loading test.scv**
```{r}
      setInternet2()
      file_test <-  "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
      download.file(file_test,"file_test.csv")
      test_data <-  read.csv("file_test.csv") ## data for prediction
      test_data <- test_data[,-1]

      test_data <- transform(test_data,"hour"=as.factor(strptime(test_data$cvtd_timestamp,
                                                                   format="%d/%m/%Y %H:%M")$hour))
      test_data <- transform(test_data,"year"=as.factor(as.Date(test_data$cvtd_timestamp,
                                                                     format="%d/%m/%Y")))
      
      test_data <- test_data[,!grepl("time",colnames(test_data))]  

      test_data <- test_data[,!grepl("window",colnames(test_data))]
```

**Computing PCs**
```{r}
      PCAtest_data_val <- predict(PCAtrain,test_data[,!na_count2&!na_count&!factorVar])
      test_data_factor_var <- test_data[,colnames(test_data)[!na_count2&!na_count&factorVar]]
      ##test_data_var <- cbind(test_data_factor_var[,c("user_name","problem_id")],PCAtest_data_val)
      test_data_var <- cbind(test_data_factor_var,PCAtest_data_val)
      colnames(test_data_var)[1] <- "user_name"
```

**Adjusting Factor values in test_data**  
In order to generate prediction the factor types need to have the same attributes (factor levels) as in the train_data sets (training, testing, validation). Otherwise the *predict()* function will not work for the models.
```{r}
      test_data_var[,2]<- factor("E",levels=c("A","B","C","D","E"))
      colnames(test_data_var)[2]<-"classe"
      
      levels(test_data_var[,4]) <- c(levels(test_data_var[,4]),levels(validation_var[,4])[!is.element(
                                              levels(validation_var[,4]),levels(test_data_var[,4]))])
      levels(test_data_var[,3]) <- c(levels(test_data_var[,3]),levels(validation_var[,3])[!is.element(
                                              levels(validation_var[,3]),levels(test_data_var[,3]))])
      levels(test_data_var[,2]) <- c(levels(test_data_var[,2]),levels(validation_var[,2])[!is.element(
                                              levels(validation_var[,2]),levels(test_data_var[,2]))])
      levels(test_data_var[,1]) <- c(levels(test_data_var[,1]),levels(validation_var[,1])[!is.element(
                                              levels(validation_var[,1]),levels(test_data_var[,1]))])
```

**Predictions for Individual Models**
```{r}
      rf_test_data  <- predict(modelFit_rf,test_data_var)
      rrf_test_data <- predict(modelFit_rrf,test_data_var)
      nb_test_data <- predict(modelFit_nb,test_data_var)
      svm_test_data  <- predict(modelFit_svm,test_data_var)
      gbm_pred_test_data <- predict(modelFit_gbm, test_data_var, n.trees=1000, type="response")
      gbm_pred_test_data <- factor(colnames(gbm_pred_test_data)[max.col(data.frame(
                              gbm_pred_test_data))],levels=c("A","B","C","D","E"))
      nnet_pred_test_data  <- predict(modelFit_nnet,test_data_var)
      nnet_pred_test_data <- factor(colnames(nnet_pred_test_data)[max.col(data.frame(
                              nnet_pred_test_data))],levels=c("A","B","C","D","E"))
```

**Prediction for Combined Model**
```{r}
      predDF_test_data <- data.frame("rf"=predict(modelFit_rf,test_data_var),"nb"=predict(modelFit_nb,test_data_var),
                          "svm"=predict(modelFit_svm,test_data_var),
                          "gbm"=gbm_pred_test_data,
                          "nnet"=nnet_pred_test_data,
                          "outc"=test_data_var$classe)
      combPred_test_data <- predict(combModFit,predDF_test_data)
```

**Summarizing the Predictions and Plotting**
```{r results = 'asis'}
      test_data_table <- data.frame(cbind("rf"=as.character(rf_test_data),"rrf"=as.character(rrf_test_data),
                                          "nb"=as.character(nb_test_data),"svm"=as.character(svm_test_data),
                                          "gbm"=as.character(gbm_pred_test_data), "nnet"=as.character(nnet_pred_test_data),
                                          "comb"=as.character(combPred_test_data)))
      rownames(test_data_table) <- paste("id",1:nrow(test_data_table),sep="_")
      html_attributes <- list('border="1" align="center"')
      print(xtable(t(test_data_table),caption="Predictions for test.csv"),
            type="html",html.table.attributes=html_attributes)
```
